---
title: "Archetype ggplot Graphs"
author: "Nick Gotelli"
date: "December 22, 2018"
output: 
  html_document: 
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file condenses and simplifies a terrific set of graphs originally posted at:
http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html#top 

To make the examples self-contained, all library calls occur within each example

# Correlation

## Simple scatterplot

```{r}
# Scatterplot

library(ggplot2)

scatter1 <- ggplot(mpg, aes(x=cty, y=hwy)) +
  geom_point(aes(col=cyl,size=displ)) +
  geom_smooth(method="loess",se=FALSE,col="red") +  
  xlim(c(0, 40)) + 
  ylim(c(0, 50)) + 
  labs(subtitle="Highway vs City Mileage", 
       y="City Mileage (mpg)", 
       x="Highway Mileage (mpg)", 
       title="Scatterplot", 
       caption = "Source: R-data") +
  theme_bw() 
  

plot(scatter1)

```

## Scatterplot with encircling

```{r}

# Scatterplot with encircling
# devtools::install_github("hrbrmstr/ggalt")
library(ggplot2)
library(ggalt)

mpgGrp <- mpg[mpg$class=="compact",] # create a subset data frame for encircling

scatter2 <- ggplot(mpg, aes(x=cty, y=hwy)) +
  geom_point(aes(col=cyl,size=displ)) +
  geom_smooth(method="loess",se=FALSE,col="red") +  
  xlim(c(0, 40)) + 
  ylim(c(0, 50)) + 
  labs(subtitle="Highway vs City Mileage", 
       y="City Mileage (mpg)", 
       x="Highway Mileage (mpg)", 
       title="Scatterplot with encircling", 
       caption = "Source: R-data") +
  theme_bw() +
  
  geom_encircle(data=mpgGrp,
                aes(x=cty,y=hwy),
                     color="seagreen",
                     size=3,
                     expand=0.08)

plot(scatter2)


```

## Jitter plot
```{r}
# Jitter plot

library(ggplot2)

scatter3 <- ggplot(mpg, aes(x=cty, y=hwy)) +
  geom_jitter(width=1, height=1, size=1) +
              # aes(col=cyl,size=displ)) +
  geom_smooth(method="loess",se=FALSE,col="red") +  
  xlim(c(0, 40)) + 
  ylim(c(0, 50)) + 
  labs(subtitle="Highway vs City Mileage", 
       y="City Mileage (mpg)", 
       x="Highway Mileage (mpg)", 
       title="Scatterplot with jittering", 
       caption = "Source: R-data") +
  theme_bw() 
  

plot(scatter3)

```

## Counts chart

```{r}
library(ggplot2)

scatter4 <- ggplot(mpg, aes(x=cty, y=hwy)) +
  geom_count(col="goldenrod",show.legend=TRUE) +
              # aes(col=cyl,size=displ)) +
  geom_smooth(method="loess",se=FALSE,col="red") +  
  xlim(c(0, 40)) + 
  ylim(c(0, 50)) + 
  labs(subtitle="Highway vs City Mileage", 
       y="City Mileage (mpg)", 
       x="Highway Mileage (mpg)", 
       title="Scatterplot with count overlay", 
       caption = "Source: R-data") +
  theme_bw() 
  

plot(scatter4)


```


## Bubble plot

```{r}
library(ggplot2)

mpg_select <- mpg[mpg$manufacturer %in% c("audi", "ford", "honda", "hyundai"), ]

scatter5 <- ggplot(mpg_select, aes(x=displ, y=cty)) +
  geom_smooth(aes(col=manufacturer),method="lm",se=FALSE) +  
  geom_jitter(aes(col=manufacturer,size=hwy)) +
  labs(subtitle="Highway Mileage vs Displacement", 
       y="City Mileage (mpg)", 
       x="Highway Mileage (mpg)", 
       title="Bubble plot with highway as point size", 
       caption = "Source: R-data") +
  theme_bw() 
  

plot(scatter5)
```

## Scatter plot with marginal distributions

```{r}
library(ggplot2)
library(ggExtra)
xVar <- rnorm(300)
yVar <- xVar + 0.5*rnorm(300)
sData <- data.frame(xVar,yVar)

scatter6 <- ggplot(sData,aes(x=xVar,y=yVar)) +
  geom_point() + 
  geom_smooth(method="lm",se=FALSE) +
  theme_bw()  

ggMarginal(p=scatter6, type = "histogram", fill="thistle")
# ggMarginal(p=scatter6, type = "boxplot", fill="khaki")
# ggMarginal(p=scatter6, type = "density", fill="tomato")
# ggMarginal(p=scatter6, type = "violin", fill="wheat")



  
```

## Correlogram

```{r}
# devtools::install_github("kassambara/ggcorrplot")
library(ggplot2)
library(ggcorrplot)

# Correlation matrix
data(mtcars)
corr <- round(cor(mtcars), 1)

# Plot
ggcorrplot(corr, hc.order = TRUE, 
           type = "lower", 
           lab = TRUE, 
           lab_size = 3, 
           method="circle", 
           colors = c("tomato2", "white", "springgreen3"), 
           title="Correlogram of mtcars", 
           ggtheme=theme_bw)
```

# Deviation

## Diverging bars

```{r}
library(ggplot2)

# Data Prep
data("mtcars")  # load data
mtcars$`car name` <- rownames(mtcars)  # create new column for car names
mtcars$mpg_z <- round((mtcars$mpg - mean(mtcars$mpg))/sd(mtcars$mpg), 2)  # compute normalized mpg
mtcars$mpg_type <- ifelse(mtcars$mpg_z < 0, "below", "above")  # above / below avg flag
mtcars <- mtcars[order(mtcars$mpg_z), ]  # sort
mtcars$`car name` <- factor(mtcars$`car name`, levels = mtcars$`car name`)  # convert to factor to retain sorted order in plot.

# Diverging Barcharts
ggplot(mtcars, aes(x=`car name`, y=mpg_z, label=mpg_z)) + 
  geom_bar(stat='identity', aes(fill=mpg_type), width=.5)  +
  scale_fill_manual(name="Mileage", 
                    labels = c("Above Average", "Below Average"), 
                    values = c("above"="#00ba38", "below"="#f8766d")) + 
  labs(subtitle="Normalised mileage from 'mtcars'", 
       title= "Diverging Bars") +
  
  coord_flip() +
  theme_bw()
```

## Diverging lollipop chart

```{r}
library(ggplot2)

ggplot(mtcars, aes(x=`car name`, y=mpg_z, label=mpg_z)) + 
  geom_point(stat='identity', fill="black", size=6)  +
  geom_segment(aes(y = 0, 
                   x = `car name`, 
                   yend = mpg_z, 
                   xend = `car name`), 
               color = "black") +
  geom_text(color="white", size=2) +
  labs(title="Diverging Lollipop Chart", 
       subtitle="Normalized mileage from 'mtcars': Lollipop") + 
  ylim(-2.5, 2.5) +
  coord_flip() +
  theme_bw()
```

## Diverging dot plot

```{r}
library(ggplot2)

# Plot
ggplot(mtcars, aes(x=`car name`, y=mpg_z, label=mpg_z)) + 
  geom_point(stat='identity', aes(col=mpg_type), size=6)  +
  scale_color_manual(name="Mileage", 
                     labels = c("Above Average", "Below Average"), 
                     values = c("above"="#00ba38", "below"="#f8766d")) + 
  geom_text(color="white", size=2) +
  labs(title="Diverging Dot Plot", 
       subtitle="Normalized mileage from 'mtcars': Dotplot") + 
  ylim(-2.5, 2.5) +
  coord_flip() +
  theme_bw()
```

## Area chart

```{r}
library(ggplot2)
library(quantmod)
data("economics", package = "ggplot2")

# Compute % Returns
economics$returns_perc <- c(0, diff(economics$psavert)/economics$psavert[-length(economics$psavert)])

# Create break points and labels for axis ticks
brks <- economics$date[seq(1, length(economics$date), 12)]
lbls <- lubridate::year(economics$date[seq(1, length(economics$date), 12)])

# Plot
ggplot(economics[1:100, ], aes(date, returns_perc)) + 
  geom_area() + 
  scale_x_date(breaks=brks, labels=lbls) + 
  theme(axis.text.x = element_text(angle=90)) + 
  labs(title="Area Chart", 
       subtitle = "Perc Returns for Personal Savings", 
       y="% Returns for Personal savings", 
       caption="Source: economics")
```

# Ranking

## Ordered Bar Chart

```{r}
# Prepare data: group mean city mileage by manufacturer.
library(ggplot2)
cty_mpg <- aggregate(mpg$cty, by=list(mpg$manufacturer), FUN=mean)  # aggregate
colnames(cty_mpg) <- c("make", "mileage")  # change column names
cty_mpg <- cty_mpg[order(cty_mpg$mileage), ]  # sort
cty_mpg$make <- factor(cty_mpg$make, levels = cty_mpg$make)  # to retain the order in plot.
head(cty_mpg, 4)

theme_set(theme_bw())

# Draw plot
ggplot(cty_mpg, aes(x=make, y=mileage)) + 
  geom_bar(stat="identity", width=.5, fill="tomato3") + 
  labs(title="Ordered Bar Chart", 
       subtitle="Make Vs Avg. Mileage", 
       caption="source: mpg") + 
  theme(axis.text.x = element_text(angle=65, vjust=0.6))
```

## Lollipop Chart

```{r}

library(ggplot2)
theme_set(theme_bw())

# Plot
ggplot(cty_mpg, aes(x=make, y=mileage)) + 
  geom_point(size=3) + 
  geom_segment(aes(x=make, 
                   xend=make, 
                   y=0, 
                   yend=mileage)) + 
  labs(title="Lollipop Chart", 
       subtitle="Make Vs Avg. Mileage", 
       caption="source: mpg") + 
  theme(axis.text.x = element_text(angle=65, vjust=0.6))

```
## Dot Plot

```{r}
library(ggplot2)
library(scales)
theme_set(theme_classic())

# Plot
ggplot(cty_mpg, aes(x=make, y=mileage)) + 
  geom_point(col="tomato2", size=3) +   # Draw points
  geom_segment(aes(x=make, 
                   xend=make, 
                   y=min(mileage), 
                   yend=max(mileage)), 
               linetype="dashed", 
               size=0.1) +   # Draw dashed lines
  labs(title="Dot Plot", 
       subtitle="Make Vs Avg. Mileage", 
       caption="source: mpg") +  
  coord_flip()
```

## Slope Chart

```{r}
library(ggplot2)
library(scales)
theme_set(theme_classic())

# prep data
df <- read.csv("https://raw.githubusercontent.com/selva86/datasets/master/gdppercap.csv")
colnames(df) <- c("continent", "1952", "1957")
left_label <- paste(df$continent, round(df$`1952`),sep=", ")
right_label <- paste(df$continent, round(df$`1957`),sep=", ")
df$class <- ifelse((df$`1957` - df$`1952`) < 0, "red", "green")

# Plot
p <- ggplot(df) + geom_segment(aes(x=1, xend=2, y=`1952`, yend=`1957`, col=class), size=.75, show.legend=F) + 
                  geom_vline(xintercept=1, linetype="dashed", size=.1) + 
                  geom_vline(xintercept=2, linetype="dashed", size=.1) +
                  scale_color_manual(labels = c("Up", "Down"), 
                                     values = c("green"="#00ba38", "red"="#f8766d")) +  # color of lines
                  labs(x="", y="Mean GdpPerCap") +  # Axis labels
                  xlim(.5, 2.5) + ylim(0,(1.1*(max(df$`1952`, df$`1957`))))  # X and Y axis limits

# Add texts
p <- p + geom_text(label=left_label, y=df$`1952`, x=rep(1, NROW(df)), hjust=1.1, size=3.5)
p <- p + geom_text(label=right_label, y=df$`1957`, x=rep(2, NROW(df)), hjust=-0.1, size=3.5)
p <- p + geom_text(label="Time 1", x=1, y=1.1*(max(df$`1952`, df$`1957`)), hjust=1.2, size=5)  # title
p <- p + geom_text(label="Time 2", x=2, y=1.1*(max(df$`1952`, df$`1957`)), hjust=-0.1, size=5)  # title

# Minify theme
p + theme(panel.background = element_blank(), 
           panel.grid = element_blank(),
           axis.ticks = element_blank(),
           axis.text.x = element_blank(),
           panel.border = element_blank(),
           plot.margin = unit(c(1,2,1,2), "cm"))
```

