---
title: "Multiple plots in `ggplot2` with `patchwork`"
author: "Nick Gotelli"
date: "February 5, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE)
```

The `patchwork` package makes it very easy to create layouts in ggplot that have multiple panels.

The package is by Thomas Lin Pederson, and you can get it from his GitHub site:

https://github.com/thomasp85/patchwork

# Installation
This is a little tricky, because the installation is not from CRAN. First, you need to install devtools, which is available from CRAN. Then load the devtools library and run the `install_github` command from the console:

```{r, eval=FALSE}
# install.packages("devtools")
library(devtools)
install_github("thomasp85/patchwork")
```

I had a bit of trouble with this and ended up re-installing ggplot, but eventually it all worked.

# Sample Data Sets
Here are some quick and dirty artificial data just to have something to plot:

```{r}
library(ggplot2)
library(patchwork)

d1 <- runif(500)
d2 <- rep(c("Treatment","Control"),each=250)
d3 <- rbeta(500,shape1=100,shape2=3)
d4 <- d3 + rnorm(500,mean=0,sd=0.1)
plotData <- data.frame(d1,d2,d3,d4)
str(plotData)
```
# Create 4 Simple `ggplot` Graphs
```{r}
p1 <- ggplot(data=plotData) + geom_point(aes(x=d3, y=d4))
p1

p2 <- ggplot(data=plotData) + geom_boxplot(aes(x=d2,y=d1,fill=d2))+
theme(legend.position="none")
p2

p3 <- ggplot(data=plotData) +
  geom_histogram(aes(x=d1, color=I("black"),fill=I("orchid")))
p3

p4 <- ggplot(data=plotData) +
  geom_histogram(aes(x=d3, color=I("black"),fill=I("goldenrod")))
p4
```

# Place Two Plots Horizontally
```{r}
p1 + p2
```

# Place Three Plots Vertically
```{r}
p1 + p2 + p3 + plot_layout(ncol=1)
```

# Change Relative Area Of Each Plot

```{r}
p1 + p2 + plot_layout(ncol=1,heights=c(2,1))

p1 + p2 + plot_layout(ncol=2,widths=c(2,1))
```

# Add A Spacer Plot
```{r}
# not working correctly!
p1 + plot_spacer() + p2
```

# Use Nested Layouts

```{r}
p1 + {
  p2 + {
    p3 +
      p4 +
      plot_layout(ncol=1)
  }
} +
  plot_layout(ncol=1)
```

# `-` Operator For Subtrack Placement

```{r}
p1 + p2 - p3 + plot_layout(ncol=1)
```

# `\` and `|` Operators For Intuitive Layouts

```{r}
(p1 | p2 | p3)/p4

(p1 | p2)/(p3 | p4)

(p1 / p2 /p3) | p4
```

These last seem the most useful and very easy to remember.

There are a few other features on the `patchwork` site to read about, but this gives us the basic functionality for creating composite panels of any kind.
